---
title: "BIMM 143 Lab 10"
author: "Ashley Martinez (PID: A17891957)"
format: pdf
toc: true
---

As it is nearly Halloween and the half way point in the quarter, let's do a mini project to help us figure out the best candy!

## Candy Data
Our data comes from the 538 website and is avaliable as a CSV file:

```{r}
candy<- read.csv("candy-data.txt",row.names=1)
flextable::flextable(head(candy,10))
```

>Q1. How many different candy types are in this dataset?

**Answer**: 85

```{r}
nrow(candy)

candy|>
  nrow()

library(tidyverse)

candy%>%
  nrow()
```


>Q2. How many fruity candy types are in the dataset?

**Answer**: 38

```{r}
sum(candy$fruity)
```

## Favorite Candy
My favorite winpercent

```{r}
candy["Twix",]$winpercent
candy["Kit Kat",]$winpercent
candy["Tootsie Roll Snack Bars",]$winpercent
```

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

**Answer**: 81.64%

>Q4. What is the winpercent value for “Kit Kat”?

**Answer**: 76.77%

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

**Answer**: 49.65%

Here is an alternative way to calculate winpercent.

```{r}
library(dplyr)

candy|>
  filter(rownames(candy)=="Twix") |>
  select(winpercent)
```

```{r}
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

**Answer**: The winpercent row seems to be on a different scale because all the other rows are on a scale from 0 to 1 while this row ranges from around 14 to 84, it is on a 0-100 scale.

>Q7. What do you think a zero and one represent for the candy$chocolate column?

**Answer**: The candy doesn't contain chocolate if it is 0.

>Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy)+
  aes(winpercent)+
  geom_histogram(bins=20)
```

>Q9. Is the distribution of winpercent values symmetrical?

**Answer**: The distribution is not symmetrical with the maximum skewed to the left.

>Q10. Is the center of the distribution above or below 50%?

**Answer**: The center is below 50%.

```{r}
mean(candy$winpercent)
summary(candy$winpercent)
```


>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

**Answer**: Chocolate is higher ranked.

To do this:

#1. Find all chocolate candy in the dataset

#2. Find their winpercent values

#3. Calculate the mean of these values

#4-6. Do the same for fruity candy

#7. Compare mean winspercents of chocolate vs. fruity

#8. Pick the highest as the winner

```{r}
choc.inds<-candy$chocolate==1
choc.win<-candy[choc.inds,]$winpercent
choc.mean<-mean(choc.win)
choc.mean

fruit.inds<-candy$fruity==1
fruit.win<-candy[fruit.inds,]$winpercent
fruit.mean<-mean(fruit.win)
fruit.mean
```

```{r}
mean(candy[candy$fruity==1,]$winpercent)
```


>Q12. Is this difference statistically significant?

**Answer**: Yes, the p value is less than 0.05.

```{r}
t.test(choc.win,fruit.win)
```

## Overall Candy Rankings

>Q13. What are the five least liked candy types in this set?

**Answer**: Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters

```{r}
candy|>
  arrange(winpercent)|>
  head(5)
```


>Q14. What are the top 5 all time favorite candy types out of this set?

**Answer**: Snickers, Kit Kat, Twix, Reese's Minatures, and Reese's Peanut Butter cup

```{r}
candy|>
  arrange(winpercent)|>
  tail(5)
```

>Q15. Make a first barplot of candy ranking based on winpercent values.

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

>Q17. What is the worst ranked chocolate candy?

**Answer**: Sixlets

>Q18. What is the best ranked fruity candy?

**Answer**: Starburst

```{r}
my_cols<-rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```

## Price Percent

```{r}
library(ggrepel)
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 7)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

**Answer**: Reese's Miniatures

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

**Answer**: Nik L Nip, Nestle Smarties, ring pop, Hershey's Krackel, Hershey's Milk Chocolate

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

## Correlation Structure

```{r}
library(corrplot)
cij<- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

**Answer**: Chocolate and fruity

>Q23. Similarly, what two variables are most positively correlated?

**Answer**: Chocolate and bar

## Principal Component Analysis

The function to use is called 'prcomp()' with an optional 'scale=T/F' argument. 

```{r}
pca<- prcomp(candy,scale=TRUE)
summary(pca)
```

Our main PCA result figure

```{r}
ggplot(pca$x)+
  aes(PC1,PC2,label=rownames(pca$x))+
  geom_point(col=my_cols)+
  geom_text(col=my_cols)
```

We should also examine the variable "loadings" or contributions of the original variable to the new PCs.

```{r}
ggplot(pca$rotation)+
  aes(PC1,rownames(pca$rotation))+
  geom_col()
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

**Answer**: Pluribus, hard, and fruity. These make sense because these tend to be the main trends for all fruity candy.

```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

Interactive plots that can be zoomed on and "brushed" over can be made with the **plotly** package. It's output is interactive and will not render to PDF.



