---
title: "Class 6: R Functions"
author: "Ashley Martinez (PID: A17891957)"
format: pdf
---

All functions in R have at least 3 things:

-A **name**, we pick this and use it to call the function.
-Input **arguments**, there can be multiple comma separated inputs to the function.
-The **body**, lines of R code that do the work of the function.

Our first function:


```{r}
add <- function(x,y=1){
  x+y
}
```

Let's test our function

```{r}
add(c(1,2,3),y=10)
```

```{r}
add(10,100)
```


##A second function

Let's try something more interesting. Make a sequence generation tool.

The 'sample()' function could be useful here. 

```{r}
sample(1:10,size=3)
```

Change this to work with the nucleotides A,C,G,T and return 3 of them.

```{r}
n<-c("A","C","G","T")
sample(n,size=15, replace=TRUE)
```

Turn this snipet into function that returns a user specified length DNA sequence. Let's call it 'generate_dna()'...

```{r}
n<-c("A","C","G","T")
sample(n,size=length(n), replace=TRUE)
```

```{r}
generate_dna<-function(len=10,fasta=FALSE){
  n<-c("A","C","G","T")
  v<-sample(n,size=len, replace=TRUE)
  
  #Make a single element vector
  s<-paste(v,collapse="")
  
cat("Well done you!\n")
if (fasta){
return(s)
} else{
  return(v)
}
}
```

```{r}
generate_dna(5)
```

```{r}
s<-generate_dna(15)
s
```

I want the option to return a single element vector with my sequence all together like this: "GGAGTAC"

```{r}
s
paste(s,collapse="")
```


```{r}
generate_dna(10,fasta=TRUE)
```

##A more advanced example

Make a third function that generates protein sequence of a user specified length and format.

```{r}
generate_protein<-function(size=15,fasta=TRUE){
  aa<-c("A","R","N","D","C","E","Q","G","H","I",
           "L","K","M","F","P","S","T","W","Y","V","U")
  seq<-sample(aa, size=size, replace=TRUE)
  
out<- if (fasta) paste(seq,collapse="") else seq
  return(out)
}
```

Try this out...

```{r}
generate_protein(10)
```

> Q. Generate random protein sequences between lengths 5 and 12 amino-acids.

```{r}
generate_protein(5)
```

One approach is to do this by brute force calling our function for each length 5 to 12. 

Another approach is to write a "for()' loop to itterate over the input values 5 to 12.

A very useful third approach that is R specific is to use the 'sapply()' function.

```{r}
seq_lengths<-6:12
for(i in seq_lengths){
  cat(">",i, "\n")
  cat(generate_protein(i))
  cat("\n")
}
```

```{r}
sapply(6:12,generate_protein)
```


> **Key Point**: Writing functions in R is doable but not the easiest thing in the world starting with our working snippet of code then using LLM tools to improve and generalize your function code is a productive approach.




